\section{Multiphysics Finite-Element Formulation}
\label{sec:multiphysics:formulation}

Within the PETSc solver framework, we want to solve a system of
partial differential equations in which the weak form can be
expressed as $F(t,s,\dot{s}) = G(t,s)$, $s(t_0) = s_0$, where $F$ and
$G$ are vector functions, $t$ is time, and $s$ is the solution vector.

Using the finite-element method we manipulate the weak form of the
system of equations involving a vector field $\vec{u}$ into integrals
over the domain $\Omega$ matching the form,
\begin{equation}
  \label{eqn:problem:form}
  \int_\Omega \trialvec[u] \cdot \vec{f}_0(t,s,\dot{s}) + \nabla \trialvec[u] : \tensor{f}
_1(t,s,\dot{s}) \, 
d\Omega =
  \int_\Omega \trialvec[u] \cdot \vec{g}_0(t,s) + \nabla \trialvec[u] : \tensor{g}_1(t,s) \, 
d\Omega,
\end{equation}
where $\trialvec[u]$ is the trial function for field $\vec{u}$,
$\vec{f}_0$ and $\vec{g}_0$ are vectors, and $\tensor{f}_1$ and
$\tensor{g}_1$ are tensors. With multiple partial differential
equations we will have multiple equations of this form, and the
solution vector $s$, which we usually write as $\vec{s}$, will be
composed of several different fields, such as displacement $\vec{u}$,
velocity $\vec{v}$, pressure $p$, and temperature $T$. Boundary
conditions will also contribute similar terms with integrals over the
corresponding boundaries.

For consistency with the PETSc time stepping formulation, we call
$G(t,s)$ the RHS function and call $F(t,s,\dot{s})$ the LHS (or I)
function. Likewise, the Jacobian of $G(t,s)$ is the RHS Jacobian and
the Jacobian of $F(t,s,\dot{s})$ is the LHS Jacobian.  Using a
finite-element discretization we break up the domain and boundary
integrals into sums over the cells and boundary faces/edges,
respectively. Using numerical quadrature those sums in turn involve
sums over the values at the quadrature points with appropriate
weights. Thus, computation of the RHS function boils down to pointwise
evaluation of $\vec{g}_0(t,s)$ and $\tensor{g}_1(t,s)$, and
computation of the LHS function boils down to pointwise evaluation of
$\vec{f}_0(t,s,\dot{s})$ and $\tensor{f}_1(t,s,\dot{s})$.

\subsection{Jacobian}

The LHS Jacobian $J_F = \frac{\partial F}{\partial s} +
s_\mathit{tshift} \frac{\partial F}{\partial \dot{s}}$ and the RHS
Jacobian $J_G = \frac{\partial G}{\partial s}$, where
$s_\mathit{tshift}$ arises from the temporal discretization. We put
the Jacobians for each equation into the form:
\begin{align}
  \label{eqn:jacobian:form}
  J_F &= \int_\Omega \trialvec \cdot \tensor{J}_{f0}(t,s,\dot{s}) \cdot \basisvec
  + \trialvec \cdot \tensor{J}_{f1}(t,s,\dot{s}) : \nabla \basisvec
  + \nabla \trialvec : \tensor{J}_{f2}(t,s,\dot{s}) \cdot \basisvec
  + \nabla \trialvec : \tensor{J}_{f3}(t,s,\dot{s}) : \nabla \basisvec \, d\Omega \\
%
  J_G &= \quad \int_\Omega \trialvec \cdot \tensor{J}_{g0}(t,s) \cdot \basisvec
  + \trialvec \cdot \tensor{J}_{g1}(t,s) : \nabla \basisvec
  + \nabla \trialvec : \tensor{J}_{g2}(t,s) \cdot \basisvec
  + \nabla \trialvec : \tensor{J}_{g3}(t,s) : \nabla \basisvec \, d\Omega,
\end{align}
where $\basisvec$ is a basis function.  Expressed in index notation
the Jacobian coupling solution field components $s_i$ and $s_j$ is
\begin{equation}
\label{eqn:jacobian:index:form}
J^{s_is_j} = \int_\Omega \trialscalar_i J_0^{s_is_j} \basisscalar_j + \trialscalar_i 
J_1^{s_js_jl} 
\basisscalar_{j,l} + \trialscalar_{i,k} J_2^{s_is_jk} \basisscalar_j + \trialscalar_{i,k} 
J_3^{s_is_jkl} 
\basisscalar_{j,l} \, d\Omega, 
\end{equation}
It is clear that the tensors $J_0$, $J_1$, $J_2$, and $J_3$ have
various sizes: $J_0(n_i,n_j)$, $J_1(n_i,n_j,d)$, $J_2(n_i,n_j,d)$,
$J_3(n_i,n_j,d,d)$, where $n_i$ is the number of components in
solution field $s_i$, $n_j$ is the number of components in solution
field $s_j$, and $d$ is the spatial dimension.  Alternatively,
expressed in discrete form, the Jacobian for the coupling between
solution fields $s_i$ and $s_j$ is
\begin{equation}
  \label{eqn:jacobian:discrete:form}
  J^{s_is_j} = J_{0}^{s_is_j} + J_{1}^{s_is_j} B + B^T J_{2}^{s_is_j} + B^T J_{3}^{s_is_j} B,
\end{equation}
where $B$ is a matrix of the derivatives of the basis functions and $B^T$
is a matrix of the derivatives of the trial functions. 

\important{See
  \url{https://www.mcs.anl.gov/petsc/petsc-master/docs/manualpages/FE/PetscFEIntegrateJacobian.html}
  for the ordering of indicies in the Jacobian pointwise functions.}

\subsection{PETSc TS Notes}

\subsubsection{Explicit Time Stepping}

Explicit time stepping with the PETSc TS requires
$F(t,s,\dot{s}) = \dot{s}$.
\begin{itemize}
\item We do not specify the functions $\vec{f}_0(t,s,\dot{s})$ and
  $\tensor{f}_1(t,s,\dot{s})$ because the PETSc TS will assume
  $F(t,s,\dot{s}) = \dot{s}$ if no LHS (or I) function is given.
\item The PETSc TS will verify that the LHS (or I) function is null.
\item We also do not specify $J_F$ or $J_G$.
\item This leaves us with only needing to specify $\vec{g}_0(t,s)$
  and $\tensor{g}_1(t,s)$. 
\end{itemize}

For explicit time stepping with the PETSc TS, we need
$F(t,s,\dot{s}) = \dot{s}$. Using a finite-element formulation for
elastodynamics, $F(t,s,\dot{s})$ generally involves integrals of the
inertia over the domain. It is tempting to simply move these terms to
the RHS, but that introduces inertial terms into the boundary
conditions, which makes them less intuitive. Instead, we transform our
equation into the form $\dot{s} = G^*(t,s)$ where
$G^*(t,s) = M^{-1} G(t,s)$. We take $M$ to be a lumped (diagonal)
matrix, so that $M^{-1}$ is trivial to compute. In computing the RHS
function, $G^*(t,s)$, we compute $G(t,s)$, then compute $M$ and
$M^{-1}$, and then $M^{-1}G(t,s)$. For the elasticity equation with
inertial terms, $M$ contains the mass matrix.

\subsubsection{Implicit Time Stepping}

The LHS (or I) function is associated with implicit time-stepping.
When using implicit time-stepping, we place all of the terms on the
LHS. Even though placing all of the terms on the LHS sometimes
requires different pointwise functions for implicit and explicit time
stepping, it minimizes the number of pointwise functions needed for
implicit time stepping.  If no RHS function is given, then the PETSc
TS assumes $G(t,s) = 0$, so we only need to specify $F(t,s,\dot{s})$
and $J_f$.

\subsubsection{Implicit-Explicit Time Stepping}

For implicit-explicit time stepping algorithms, the equations
integrated with explicit time stepping have $\dot{s}$ as the LHS
function, and the equations integrated with implicit time stepping
have 0 as the RHS function.

