\chapter{Extending PyLith}
\label{cha:extending}

One of the powerful features of using the Pyre framework in PyLith is
the ability to extend the functionality of the software without
altering any of the PyLith code. Any of the components can be replaced
with other compatible components. You are already familiar with this
feature from running the examples; when you set the spatial database
to \object{UniformDB}, \object{SimpleDB}, or \object{SCECCVMH} you are
switching between different compatible components for a spatial
database facility. Modifying the governing equations to include other
physical processes requires changing the data structures associated
with the solution and altering the PyLith code.

In this section we provide examples of how to extend PyLith for components
that users will most likely want to replace with their own custom
versions. You will need a familiarity with Python, Makefiles, and
C++ to write your own components. The primary steps in constructing
a component to extend PyLith's functionality include:
\begin{enumerate}
\item Setting up the source files for the component or set of components
  based on the templates.
\item Edit the Python source file (\filename{py}) for the component.
  \begin{enumerate}
  \item Define the user-specified properties and facilities.
  \item Transfer the user-specified data from the Python object to the corresponding
    C++ object via calls to the SWIG interface object.
  \end{enumerate}
\item Edit the C++ header (\filename{hh}) and implementation files (\filename{cc})
  for the component.
  \begin{enumerate}
  \item Implement the methods required to satisfy the interface definition
    of the component.
  \item Implement the desired functionality of the component in C++.
  \end{enumerate}
\item Edit the SWIG interface files (\texttt{.i}) that provide the glue
  between Python and C++.
\item Edit the Python source file that tests the functionality of the component.
\item Run configure, build, install, and run the tests of the component.
\end{enumerate}

\section{Spatial Databases}
\label{sec:extending:spatialdata:databases}

PyLith provides several types of spatial databases that can be used
for specification of parameters associated with boundary conditions,
earthquake ruptures, and physical properties. In this example we demonstrate
how to provide a spatial database, UniformVelModel, for specifying
elastic properties. The source files are included in the source for
the spatialdata package in the \filename{templates/spatialdb} directory.
The \filename{README} file in \filename{templates/spatialdb} provides
detailed instructions for the various steps involved, and the source
files contain numerous comments to help guide you through the customization
process. 

The \object{UniformVelModel} component provides uniform physical
properties: P-wave speed, S-wave speed, and density. Although this is
a rather trivial specification of physical properties that could be
easily done using a UniformDB, this example demonstrates how to create
a user-defined component that matches the requirements of a spatial
database for elastic physical properties. Adding additional physical
properties is simply a matter of including some additional values in
the spatial database. Furthermore, in cases where we are constructing
a spatial database for a seismic velocity model, the data points are
georeferenced. With our uniform physical properties we do not need to
worry about any differences in coordinate systems between our seismic
velocity model and points at which the model is queried. However, in
many cases we do, so we illustrate this functionality by using a
geographic projection as the coordinate system in our example.

Using a top-down approach, the first step is to determine what
information the user will need to supply to the component. Is the data
for the spatial database in a file or a series of files? If so, file
names and possible paths to a directory containing files with known
names might be necessary. Are there other parameters that control the
behavior of the component, such as a minimum shear wave speed? In our
example the user supplies values for the P-wave speed, S-wave speed,
and density.  The user-supplied parameters become Pyre properties and
facilities in the Python source file. Because our user supplied
parameters are floating point values with dimensions, we create
dimensional properties vs, vp, and density. In addition to defining the properties of the
component, we also need to transfer these properties to the C++ object
that does the real work. This is done by calling the C++
\texttt{vs()}, \texttt{vp()}, and \texttt{density()} accessor
functions that are accessible via the Python module created by SWIG.

In the C++ object we must implement the functions that are required
by the spatial database interface. These functions are listed near
the beginning of the UniformVelModel class definition at the top of
the C++ header file, \filename{UniformVelModel.hh}. The C++ object also
includes the accessor functions that allow us to set the P-wave speed,
S-wave speed, and density values to the user-specified values in the
Python object. Additional information, such as a file name, parameters
defined as data structures, etc., would be set via similar accessor
functions. You can also add additional functions and data structures
to the C++ class to provide the necessary operations and functionality
of the spatial database. 

In \object{SimpleDB} we use a separate class to read in the spatial
database and yet another class to perform the actual query. In our
example, the C++ object also creates and stores the UTM zone 10
geographic projection for the seismic velocity model. When the spatial
database gets a query for physical properties, we transform the
coordinates of the query point from its coordinate system to the
coordinate system of our seismic velocity model.

In order to use SWIG to create the Python module that allows us to
call C++ from Python, we use a ``main'' SWIG interface file
(\filename{spatialdbcontrib.i} in this case) and then one for each
object (\filename{UniformVelModel.i} in this case). This greatly
simplifies keeping the Python module synchronized with the C++ and
Python code. The \filename{UniformVelModel.i} SWIG file is nearly
identical to the corresponding C++ header file.  There are a few
differences, as noted in the comments within the file.  LICENSE.md and
pasting the C++ header file and then doing a little cleanup is a very
quick and easy way to construct a SWIG interface file for a C++
object. Because very little changes from SWIG module to SWIG module,
it is usually easiest to construct the ``main'' SWIG interface by
LICENSE.md and customizing an existing one.

Once the Python, C++, and SWIG interface files are complete, we are
ready to build the module. The \filename{Makefile.am} file defines how
to compile and link the C++ library and generate the Python module via
SWIG. The \filename{configure.ac} file contains the information used
to build a configure script. The configure script checks to make sure
it can find all of the tools needed to build the component (C++
compiler, Python, installed spatial database package, etc.). See the
README file for detailed instructions on how to generate the configure
script, and build and install the component.

We recommend constructing tests of the component to insure that it is
functioning properly before attempting to use it in an application.
The \filename{tests} directory within \filename{templates/spatialdb}
contains a Python script, \filename{testcontrib.py}, that runs the
tests of the UniformVelModel component defined in
\filename{TestUniformVelModel.py}.  Normally, one would want to test
each function individually to isolate errors and create C++ tests as
well as the Python tests included here.  In our rather simple example,
we simply test the overall functionality of the component. For
examples of thorough testing, see the spatialdata and PyLith source
code.

Once you have built, installed, and tested the UniformVelModel, it
is time to use it in a simple example. Because the seismic velocity
model uses georeferenced coordinates, our example must also use georeferenced
coordinates. The dislocation example in the PyLith \filename{examples/twocells/twotet4-geoproj}
directory uses UTM zone 11 coordinates. The spatial database package
will transform the coordinates between the two projections as defined
in the UniformVelModel \texttt{query()} function. The dislocation
example uses the SCEC CVM-H seismic velocity model. In order to replace
the SCEC CVM-H seismic velocity with our uniform velocity model, in
\filename{pylithapp.cfg}
\begin{cfg}
# Replace these two lines
<f>db_properties</f> = spatialdata.spatialdb.SCECCVMH
<p>db_properties.data_dir</p> = /home/john/data/sceccvm-h/vx53/bin
# with
<f>db_properties</f> = spatialdata.spatialdb.contrib.UniformVelModel
\end{cfg}
When you run the dislocation example, the \filename{dislocation-statevars\_info.vtk}
file should reflect the use of physical properties from the uniform
seismic velocity with $\mu=1.69\times10^{10}\mathrm{Pa}$, $\lambda=1.6825\times10^{10}\mathrm{Pa}$,
and $\rho=2500\mathrm{kg/m^{3}}$.


\section{Bulk Constitutive Models}
\label{sec:extending:materials}

PyLith includes several linearly elastic and inelastic bulk
constitutive models for 2D and 3D problems. In this example, we
demonstrate how to extend PyLith by adding your own bulk constitutive
model. We reimplement the 2D plane strain constitutive model while
adding the current strain and stress tensors as state variables. This
constitutive model, \object{PlaneStrainState}, is not particularly
useful, but it illustrates the basic steps involved in creating a bulk
constitutive model with state variables. The source files are included
with the main PyLith source code in the \filename{templates/materials}
directory. The \filename{README} file in
\filename{templates/materials} provides detailed instructions for the
various steps, and the source files contain numerous comments to guide
you through the customization process.

In contrast to our previous example of creating a customized spatial
database, which involved gathering user-specified parameters via the
Pyre framework, there are no user-defined parameters for bulk
constitutive models. The specification of the physical properties and
state variables associated with the constitutive model is handled
directly in the C++ code. As a result, the Python object for the
constitutive model component is very simple, and customization is
limited to simply changing the names of objects and labels.

The properties and state variables used in the bulk constitutive model
are set using arguments to the constructor of the C++ \object{ElasticMaterial}
object. We define a number of constants at the top of the C++ file
and use the \object{Metadata} object to define the properties and
state variables. The C++ object for the bulk constitutive component
includes a number of functions that implement elasticity behavior
of a bulk material as well as several utility routines:
\begin{description}
\item [{\texttt{\_dbToProperties()}}] Computes the physical properties
  used in the constitutive model equations from the physical properties
  supplied in spatial databases.
\item [{\texttt{\_nondimProperties()}}] Nondimensionalizes the physical
  properties used in the constitutive model equations.
\item [{\texttt{\_dimProperties()}}] Dimensionalizes the physical properties
  used in the constitutive model equations.
\item [{\texttt{\_stableTimeStepImplicit()}}] Computes the stable time
  step for implicit time stepping in quasistatic simulations given
  the current state (strain, stress, and state variables).
\item [{\texttt{\_calcDensity()}}] Computes the density given the physical
  properties and state variables. In most cases density is a physical
  property used in the constitutive equations, so this is a trivial
  function in those cases.
\item [{\texttt{\_calcStress()}}] Computes the stress tensor given the
  physical properties, state variables, total strain, initial stress,
  and initial strain.
\item [{\texttt{\_calcElasticConsts()}}] Computes the elastic constants
  given the physical properties, state variables, total strain, initial
  stress, and initial strain.
\item [{\texttt{\_updateStateVars()}}] Updates the state variables given
  the physical properties, total strain, initial stress, and initial
  strain. If a constitutive model does not use state variables, then
  this routine is omitted.
\end{description}
Because it is sometimes convenient to supply physical properties for
a bulk constitutive model that are equivalent but different from the
ones that appear in the constitutive equations (e.g., P-wave speed,
S-wave speed, and density rather then Lame's constants $\mu$, $\lambda,$
and density), each bulk constitutive model component has routines
to convert the physical property parameters and state variables a
user specifies via spatial databases to the physical property properties
and state variables used in the constitutive model equations. In quasistatic
problems, PyLith computes an elastic solution for the first time step
($-\Delta t$ to $t$). This means that for inelastic materials, we
supply two sets of functions for the constitutive behavior: one set
for the initial elastic step and one set for the remainder of the
simulation. See the source code for the inelastic materials in PyLith
for an illustration of an efficient mechanism for doing this.

The SWIG interface files for a bulk constitutive component are set up
in the same manner as in the previous example of creating a customized
spatial database component. The ``main'' SWIG interface file
(\filename{materialscontrib.i} in this case) sets up the Python
module, and the SWIG interface file for the component
(\filename{PlaneStrainState.i} in this case) defines the functions
that should be included in the Python module. Note that because the
C++ \object{ElasticMaterial} object defines a number of pure virtual
methods (which merely specify the interface for the functions and do
not implement default behavior), we must include many protected
functions in the SWIG interface file. If these are omitted, SWIG will
give a warning indicating that some of the functions remain abstract
(i.e., some pure virtual functions defined in the parent class
\object{ElasticMaterial} were not implemented in the child class
\object{PlaneStrainState}), and no constructor is created. When this
happens, you cannot create a \object{PlaneStrainState} Python object.

Once the Python, C++, and SWIG interface files are complete, you are
ready to configure and build the C++ library and Python module for the
component. Edit the \filename{Makefile.am} file as necessary, then
generate the configure script, run \filename{configure}, and then
build and install the library and module (see the \filename{README}
file for detailed instructions).

Because most functionality of the bulk constitutive model component is
at the C++ level, properly constructed unit tests for the
\object{component} should include tests for both the C++ code and
Python code. The C++ unit tests can be quite complex, and it is best
to examine those used for the bulk constitutive models included with
PyLith. In this example we create the Python unit tests to verify that
we can create a \object{PlaneStrainState} Python object and call some
of the simple underlying C++ functions.  The source files are in the
\filename{templates/materials/tests} directory.  The
\filename{testcontrib.py} Python script runs the tests defined in
\filename{TestPlaneStrainState.py}.

Once you have built, installed, and tested the \object{PlaneStrainState}
component, it is time to use it in a simple example. You can try using
it in any of the 2D examples. For the examples in \filename{examples/twocells/twoquad4},
in \filename{pylithapp.cfg}
\begin{cfg}
# Replace
<f>material</f> = pylith.materials.ElasticPlaneStrain
# with
<f>material</f> = pylith.materials.contrib.PlaneStrainState
\end{cfg}
or simply add the command line argument
\commandline{-{}-timedependent.homogeneous.material=pylith.materials.contrib.PlaneStrainState}.
when running any of the 2D examples. The output should remain
identical, but you should see the \object{PlaneStrainState} object
listed in the information written to stdout.


\section{Fault Constitutive Models}
\label{sec:extending:fault}

PyLith includes two of the most widely used fault constitutive models,
but there are a wide range of models that have been proposed to explain
earthquake source processes. In this example, we demonstrate how to
extend PyLith by adding your own fault constitutive model. We implement
a linear viscous fault constitutive model wherein the perturbation
in the coefficient of friction is linearly proportional to the slip
rate. This constitutive model, \object{ViscousFriction}, is not particularly
useful, but it illustrates the basic steps involved in creating a
fault constitutive model. The source files are included with the main
PyLith source code in the \filename{templates/friction} directory. The
\filename{README} file in \filename{templates/friction} provides detailed
instructions for the various steps, and the source files contain numerous
comments to guide you through the customization process.

Similar to our previous example of creating a customized bulk constitutive
model, the parameters are defined in the C++ code, not in the Pyre
framework. As a result, the Python object for the fault constitutive
model component is very simple and customization is limited to simply
changing the names of objects and labels.

The properties and state variables used in the fault constitutive
model are set using arguments to the constructor of the C++ \object{FrictionModel}
object, analogous to the \object{ElasticMaterial} object for bulk
constitutive models. In fact, both types of constitutive models used
the same underlying C++ object (\object{Metadata::ParamDescription})
to store the description of the parameters and state variables. We
define a number of constants at the top of the C++ file and use the
\object{Metadata} object to define the properties and state variables.
The C++ object for the fault constitutive component includes a number
of functions that implement friction as well as several utility routines:
\begin{description}
\item [{\texttt{\_dbToProperties()}}] Computes the physical properties
  used in the constitutive model equations from the physical properties
  supplied in spatial databases.
\item [{\texttt{\_nondimProperties()}}] Nondimensionalizes the physical
  properties used in the constitutive model equations.
\item [{\texttt{\_dimProperties()}}] Dimensionalizes the physical properties
  used in the constitutive model equations.
\item [{\texttt{\_dbToStateVars()}}] Computes the initial state variables
  used in the constitutive model equations from the initial values supplied
  in spatial databases.
\item [{\texttt{\_nondimStateVars()}}] Nondimensionalizes the state variables
  used in the constitutive model equations.
\item [{\texttt{\_dimStateVars()}}] Dimensionalizes the state variables
  used in the constitutive model equations.
\item [{\texttt{\_calcFriction()}}] Computes the friction stress given
  the physical properties, state variables, slip, slip rate, and normal
  traction.
\item [{\texttt{\_updateStateVars()}}] Updates the state variables given
  the physical properties, slip, slip rate, and normal traction.
\end{description}
If a constitutive model does not use state variables, then the state
variable routines are omitted. 

Because it is sometimes convenient to supply physical properties for a
fault constitutive model that are equivalent but different from the
ones that appear in the constitutive equations, each fault
constitutive model component has routines to convert the physical
property parameters and state variables a user specifies via spatial
databases to the physical property properties and state variables used
in the constitutive model equations.

The SWIG interface files for a fault constitutive component are set up
in the same manner as in the previous examples of creating a bulk
constitutive model or a customized spatial database component. The
``main'' SWIG interface file (\filename{frictioncontrib.i} in this
case) sets up the Python module, and the SWIG interface file for the
component (\filename{ViscousFriction.i} in this case) defines the
functions that should be included in the Python module. Note that
because the C++ \object{FrictionModel} object defines a number of pure
virtual methods (which merely specify the interface for the functions
and do not implement default behavior), we must include many protected
functions in the SWIG interface file. If these are omitted, SWIG will
give a warning indicating that some of the functions remain abstract
(i.e., some pure virtual functions defined in the parent class
\object{FrictionModel} were not implemented in the child class
\object{ViscousFriction}), and no constructor is created. When this
happens, you cannot create a \object{ViscousFriction} Python object.

Once the Python, C++, and SWIG interface files are complete, you are
ready to configure and build the C++ library and Python module for the
component. Edit the \filename{Makefile.am} file as necessary, then
generate the configure script, run \filename{configure}, and then
build and install the library and module (see the \filename{README}
file for detailed instructions).

Because most functionality of the fault constitutive model component
is at the C++ level, properly constructed unit tests for the
\object{component} should include tests for both the C++ code and
Python code. The C++ unit tests can be quite complex, and it is best
to examine those used for the fault constitutive models included with
PyLith. In this example we create the Python unit tests to verify that
we can create a \object{ViscousFriction} Python object and call some
of the simple underlying C++ functions.  The source files are in the
\filename{templates/friction/tests} directory.  The
\filename{testcontrib.py} Python script runs the tests defined in
\filename{TestViscousFriction.py}.

Once you have built, installed, and tested the \object{ViscousFriction}
component, it is time to use it in a simple example. You can try using
it in any of the 2D or 3D examples. For the examples in \filename{examples/bar\_shearwave/quad4,}
in \filename{shearwave\_staticfriction.cfg}
\begin{cfg}
# Replace
<f>friction</f> = pylith.friction.StaticFriction
# with
<f>friction</f> = pylith.friction.contrib.ViscousFriction
\end{cfg}
or simply add the command line argument
\commandline{-{}-timedependent.interfaces.fault.friction=pylith.friction.contrib.ViscousFriction}
when running any of the friction examples. You will also need to
supply a corresponding spatial database with the physical properties
for the viscous friction constitutive model.
