\chapter{Examples}
\label{cha:examples}

\section{Overview}

This chapter includes several suites of examples. Each suite includes
several ``steps'' which are examples that increase in complexity from
one ``step'' to the next. In some cases, a later step may make use of
output from an earlier step; these cases are clearly
documented. Table~\ref{tab:examples:overview} classifies the level of
difficulty of each example suite and provides a general description of
the type of problems discussed.

\begin{table}[htbp]
\caption{Overview of example suites.}
\label{tab:examples:overview}
\begin{tabular}{lccp{4in}}
\textbf{Directory} & \textbf{Section(s)} & \textbf{Difficulty} & \textbf{Description} \\
\hline 
\filename{twocells} & \ref{sec:example:twotri3}--\ref{sec:examples:twotet4-geoproj} & novice & Toy problems with ASCII two-cell meshes. \\
\filename{3d/hex8} & \ref{sec:example:3dhex8} & beginner & Illustration of most features using simple CUBIT box mesh. \\
\filename{3d/tet4} & \ref{sec:example:3dtet4} & beginner & Illustration of refinement using simple LaGriT box mesh. \\
\filename{bar\_shearwave} & \ref{sec:example:shearwave:tri3}--\ref{sec:example:shearwave:hex8} & beginner & Illustration of wave propagation using simple shear beam. \\
\filename{2d/subduction} & \ref{sec:example:subduction:2d} & intermediate & Illustration of coseismic, postseismic, and creep deformation using a 2-D subduction zone cross-section with a CUBIT mesh. \\
\filename{2d/greensfns} & \ref{sec:example:greensfns2d} & intermediate & Illustration of computing static Green's functions for a strike-slip and reverse fault using a CUBIT mesh. \\
\filename{3d/subduction} & \ref{sec:example:subduction:3d} & intermediate & Illustration of most PyLith features for quasi-static deformation using a 3-D subduction zone with a CUBIT mesh. \\
\hline 
\end{tabular}
The \filename{3d/subduction} example suite is the newest and most
comprehensive. Users wanting to use PyLith in their research should
work through relevant beginner examples and then the
\filename{3d/subduction} examples.
\end{table}

\subsection{Prerequisites}

Before you begin any of the examples, you will need to install PyLith
following the instructions in Chapter~\vref{cha:installation}.  For
more complex examples, you will also need either Trelis (available
from \url{csimsoft.com}), CUBIT (available to US federal government
agencies from \url{cubit.sandia.gov}) or LaGriT (available form
\url{meshing.lanl.gov}) mesh generation software to create the
meshes. If you do not wish to create your own mesh at this time, the
meshes are also provided as part of the example. The ParaView
\url{www.paraview.org} visualization package may be used to view
simulation results. ParaView 3 includes built-in documentation that is
accessed by clicking on the Help menu item. Some additional
documentation is available on the ParaView Wiki site
\url{paraview.org/Wiki/ParaView}.  You may use other visualization
software, but some adaption from what is described here will be
necessary. Furthermore, you can complete a subset of the example using
files provided (as described below), skipping the steps for which you
do not have the proper software packages installed.


\subsection{Input Files}

The files needed to work through the examples are found in the
\filename{examples} directory under the top-level PyLith
directory. There are five examples in \filename{examples/twocells},
each consisting of just two cells (elements).  These very simple
examples make use of PyLith mesh ASCII format to define the mesh. This
format is useful for understanding the basics of how PyLith works,
since it is easy to create these files by hand.  More complex
problems, such as those found in \filename{examples/3d}, use external
mesh generation software to create the meshes. All of the files used
in the example problems are extensively documented with comments.

\section{ParaView Python Scripts}
\label{sec:ParaView:Python:scripts}
\newfeature{v2.2.1}

In some of the examples (currently only the 2D and 3D subduction zone
examples) we provide ParaView Python scripts for visualizing the input
finite-element mesh and the PyLith simulation results. Some of these
scripts are very generic and are easily reused; others are more
specific to the examples. The primary advantage of the ParaView Python
scripts is that they make it easy to replicate visualizations, whether
they are produced by the developers and regenerated by users.

There are several different ways to run the ParaView Python scripts:
\begin{itemize}
\item Within the ParaView GUI, select
  \menu{Tools}$\rightarrow$\menu{Python Shell}. Override the default
  parameters as desired (which we will discuss later in this
  section). Click on the \menu{Run Script} button, and navigate to the
  select the script you want to run.
\item From a shell (terminal window) start ParaView from the command
  line with the \filename{-{}-script=FILENAME} where
  \filename{FILENAME} is the relative or absolute path to the ParaView
  Python script. Note that this method does not provide a mechanism
  for overriding the default parameters.
\item Run the ParaView Python script directly from a shell (terminal
  window) via the command line. You can use command line arguments to
  override the default values for the parameters. If pvpython is not
  in your PATH, then you can run a script called
  \filename{MY\_SCRIPT.py} using:
  \filename{PATH\_TO\_PVPYTHON/pvpython MY\_SCRIPT.py}
\end{itemize}

\tip{Running the ParaView Python script from within the ParaView GUI
  allows further manipulation of the data, which is not possible when
  running the ParaView Python script outside the ParaView GUI. When
  run outside the ParaView GUI, the interaction is limited to
  rotating, translating, and zooming.}

\important{The ParaView Python scripts run Python via
  \filename{pvpython}, which is a customized version of the Python
  interpreter included in the ParaView distribution. This is different
  from Python provided with your operating system and/or the one
  included in the PyLith distribution. This means you cannot, in
  general, import Python modules provided with the PyLith distribution
  into ParaView.}

\tip{In creating the ParaView Python scripts, we performed the steps
  within the GUI while capturing the commands using
  \menu{Tools}$\rightarrow$\menu{Start Trace} and then
  \menu{Tools}$\rightarrow$\menu{Start Trace}. This makes it very easy
  to create the Python script. Note that we have omitted supefluous
  commands in the trace when transferring the trace into a Python
  script. See the ParaView documentation for additional information
  about the Python API.}

\subsection{Overriding Default Parameters}

We setup the ParaView Python scripts, so that when they are run from
the command line in the main directory for a given example, e.g.,
\filename{examples/3d/subduction}, the script will produce the output
discussed in the manual. If you start ParaView from the OS X Dock or a
similar method, like a shortcut, then you will need to override at
least the default values for the data file(s).

In order to override the default values from within the ParaView GUI,
simply set the values within the Python shell. For example, to set the
value of the variable \object{EXODUS\_FILE} to the absolute path of
the input file,
\begin{python}[ParaView Python shell]
>>> EXODUS_FILE = "/home/johndoe/pylith/examples/3d/subduction/mesh/mesh_tet.exo"
\end{python}
In this case, we use the Python os module to get the absolute path of
the home directory and append the path to the Exodus file with the
appropriate separators for the operating system.

\important{In each of the ParaView Python scripts, the names of the
  variables and their default values are given by the DEFAULTS
  dictionary near the top of the file.}


% ======================================================================
\input{./examples/2d_box.tex}
\input{./examples/3d_box.tex}
\input{./examples/2d_strikeslip.tex}
\input{./examples/2d_reverse.tex}
\input{./examples/2d_subduction.tex}
%\input{./examples/3d_strikeslip.tex}
\input{./examples/3d_subduction.tex}

%\input{./examples/shearwave.tex}


\section{Additional Examples}


\subsection{CUBIT Meshing Examples}

The directory \filename{examples/meshing} contains several examples
of using CUBIT to construct finite-element meshes for complex geometry.
This includes features such as constructing nonplanar fault geometry
from contours, constructing topography from a DEM, and merging sheet
bodies (surfaces). A separate examples discusses defining the discretization
size using a vertex field in an Exodus-II file. See the \filename{README}
files in the subdirectories for more detailed descriptions of these
examples.


\subsection{Troubleshooting Examples}
\label{sub:troubleshooting:examples}

The directory \filename{examples/troubleshooting} contains a few examples to
practice troubleshooting a variety of user errors in parameters files and
problem setup. The files with the errors corrected are in
\filename{examples/troubleshooting/correct}.  Step-by-step corrections are
discussed in the troubleshooting PyLith simulations sessions of the
2014, 2015, 2017, and 2019 PyLith tutorials
(\url{wiki.geodynamics.org/software:pylith:start}).


\subsection{Code Verification Benchmarks}

The CIG GitHub software repository \url{https://github.com/geodynamics/pylith_benchmarks}
contains input files for a number of community benchmarks. The benchmarks
do not include the mesh files because they are so large; instead they
include the CUBIT journal files that can be used to generate the meshes.
Most, but not all, of the input files in the repository are updated
for PyLith v2.0.0, so you will need to modify them if you use another
version of PyLith.

% End of file
