
\chapter{\label{cha:Glossary}Glossary}


\section{Pyre}
\begin{description}
\item [{component}] Basic building block of a Pyre application. A component
may be built-up from smaller building blocks, where simple data types
are called properties and data structures and objects are called facilities.
In general a component is a specific implementation of the functionality
of a facility. For example, SimpleDB is a specific implementation
of the spatial database facility. A component is generally composed
of a Python object and a C++ object, although either one may be missing.
We nearly always use the naming convention such that for an object
called Foo the Python object is in file Foo.py, the C++ class definition
is in Foo.hh, inline C++ functions are in foo.icc, the C++ class implementation
is in Foo.cc, and the SWIG interface file that glues the C++ and Python
code together is in Foo.i.
\item [{facility}] Complex data type (object or data structure) building
block of a component. See component.
\item [{property}] Simple data type (string, integer, real number, or boolean
value) parameter for a component.
\end{description}

\section{DMPlex}
\begin{description}
\item [{DMPlex}] The plex construction is a representation of the topology
of the finite element mesh based upon a covering relation. For example,
segments are covered by their endpoints, faces by their bounding edges,
etc. Geometry is absent from the plex, and is represented instead
by a  field with the coordinates of the vertices. Meshes can also
be understood as directed acyclic graphs, where we call the constituents
points and arrows.
\item [{mesh}] A finite element mesh, used to partition space and provide
support for the basis functions.
\item [{cell}] The highest dimensional elements of a mesh, or mesh entities
of codimension zero.
\item [{vertex}] The zero dimensional mesh elements.
\item [{face}] Mesh elements that separate cells, or mesh entities of codimension
one.
\item [{field}] A parallel section which can be completed, or made consistent,
across process boundaries. These are used to represent continuum fields.
\item [{section}] These objects associate values in vectors to points (vertices,
 edges, faces, and cells) in a mesh. The section describes the offset
 into the vector along with the number of values associated with each
point.
\item [{dimension}] The topological dimension of the mesh, meaning the
cell dimension. It can also mean the dimension of the space in which
the mesh is embedded, but this is properly the embedding dimension.
\item [{fiber\ dimension}] Dimension of the space associated with the
field. For example, the scalar field has a fiber dimension of 1 and
a vector field has a fiber displacement equal to the dimension of
the mesh.
\item [{cohesive\ cell}] A zero volume cell inserted between any two cells
which shared a fault face. They are prisms with a fault face as the
base.
\item [{cone}] The set of entities which cover any entity in a mesh. For
example, the cone of a triangle is its three edges.
\item [{support}] The set of mesh entities which are covered by any entity
in a mesh. For example, the support of a triangle is the two tetrahedra
it separates.
\end{description}

