% -*- LaTeX -*-
%
% ----------------------------------------------------------------------
%
% Brad T. Aagaard, U.S. Geological Survey
% Charles A. Williams, GNS Science
% Matthew G. Knepley, University of Chicago
%
% This code was developed as part of the Computational Infrastructure
% for Geodynamics (http://geodynamics.org).
%
% Copyright (c) 2010-2017 University of California, Davis
%
% See COPYING for license information.
%
% ----------------------------------------------------------------------
%
\documentclass{pylithdoc}

\usepackage{units} % \nicefrac
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{array}
\usepackage{setspace} % remove

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\newcommand{\noun}[1]{\textsc{#1}}
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newenvironment{lyxcode}
{\par\begin{list}{}{
\setlength{\rightmargin}{\leftmargin}
\setlength{\listparindent}{0pt}% needed for AMS classes
\raggedright
\setlength{\itemsep}{0pt}
\setlength{\parsep}{0pt}
\normalfont\ttfamily}%
 \item[]}
{\end{list}}

\makeatother

\title{PyLith User Manual}
\author{\copyright University of California, Davis\\ Version 2.2.0}
\date{\today}

\begin{document}

\input{cover/cover.tex}
\maketitle

\frontmatter
\tableofcontents{}
\listoffigures
\listoftables

\input{preface.tex}
\mainmatter\raggedbottom

\input{./intro/intro.tex}
\input{./governingeqns/governingeqns.tex}
\input{./install/install.tex}
\input{./runpylith/runpylith.tex}
\input{./materials/materials.tex}

\input{./boundaryconditions/boundaryconditions}
\input{./examples/examples.tex}
\input{./benchmarks/benchmarks.tex}
\input{./extending/extending.tex}

\appendix
%dummy comment inserted by tex2lyx to ensure that this paragraph is not empty
\input{glossary.tex}
\input{components.tex}
\input{fileformats/fileformats.tex}
\input{materials/altformulations}
\input{analyticalsolns/analyticalsolns.tex}
\input{license.tex}

\begin{thebibliography}{Bibliography}
\bibitem[Aagaard et al., 2001a]{Aagaard:etal:2001a}Aagaard, B.T.,
J.F. Hall, and T.H. Heaton (2001), Characterization of near-source
ground motions with earthquake simulations, \emph{Earthquake Spectra,
17}(2), 177-207.

\bibitem[Aagaard et al., 2001b]{Aagaard:etal:2001b}Aagaard, B.T.,
T.H. Heaton, and J.F. Hall (2001), Dynamic earthquake ruptures in
the presence of lithostatic normal stresses: Implications for friction
models and heat production, \emph{Bulletin of the Seismological Society
of America, 91}(6), 1765-1796.

\bibitem[Aagaard et al., 2007]{Aagaard:etal:2007}Aagaard, B., C.
Williams, and M. Knepley (2007), PyLith: A finite-element code for
modeling quasi-static and dynamic crustal deformation, \emph{Eos Trans.
AGU, 88}(52), Fall Meet. Suppl., Abstract T21B-0592.

\bibitem[Aagaard et al., 2008]{Aagaard:etal:2008}Aagaard, B., C.
Williams, and M. Knepley (2008), PyLith: A finite-element code for
modeling quasi-static and dynamic crustal deformation, \emph{Eos Trans.
AGU}, \emph{89(}53), Fall Meet. Suppl., Abstract T41A-1925.

\bibitem[Bathe, 1995]{Bathe:1995}Bathe, K.-J. (1995), \textit{Finite-Element
Procedures}, Prentice Hall, Upper Saddle River, New Jersey, 1037 pp.

\bibitem[Ben-Zion and Rice, 1997]{BenZion:Rice:1997}Ben-Zion, Y.
and J.R. Rice (1997), Dynamic simulations of slip on a smooth fault
in an elastic solid, \emph{Journal of Geophysical Research}\textit{,
102}, 17,771\textendash{}17,784.

\bibitem[Brune, 1970]{Brune:1970}Brune, J.N. (1970), Tectonic stress
and spectra of seismic shear waves from earthquakes, \emph{Journal
of Geophysical Research, 75}, 4997-5009.

\bibitem[Courant et al., 1967]{Courant:etal:1967}Courant, R., K.
Friedrichs and H. Lewy (1967), On the Partial Differential Equations
of Mathematical Physics, \textit{IBM Journal of Research and Development},
11(2), 215--234.

\bibitem[Day and Ely, 2002]{Day:Ely:2002}Day, S.M. and G.P. Ely (2002),
Effct of a shallow weak zone on fault rupture: Numerical simulation
of scale-model experiments, \textit{Bull. Seismol. Soc. Am.}, 92(8),
3022-3041, doi: 10.1785/0120010273.

\bibitem[Drucker and Prager, 1952]{Drucker:Prager:1952}Drucker, D.
C. and Prager, W. (1952). Soil mechanics and plastic analysis for
limit design, \textit{Quarterly of Applied Mathematics}, \textit{10},
157\textendash{}165.

\bibitem[Liu et al., 2006]{Liu:etal:2006}Liu, P., R.J. Archuleta,
S.H. Hartzell (2006), Prediction of broadband ground-motion time histories:
Hybrid low/high-frequency method with correlated random source parameters,
\textit{Bull. Seismol. Soc. Am., 96}, 2118-2130.

\bibitem[Kaneko et al., 2008]{Kaneko:etal:2008}Kaneko, Y., N. Lapusta,
and J.-P. Ampuero (2008), Spectral element modeling of spontaneous
earthquake rupture on rate and state faults: Effect of velocity-strengthening
friction at shallow depths, \textit{Journal of Geophysical Research},\textit{
113}, B09317, doi:10.1029/2007JB005553.

\bibitem[Kaus et al., 2010]{Kaus:etal:2010}Kaus, B. J. P., H. Muhlhaus,
and D. A. May (2010), A stabilization algorithm for geodynamic numerical
simulations with a free surface, \textit{Physics of the Earth and
Planetary Interiors}, \textit{181}, 12-20, doi:10.1016/j.pepi.2010.04.007.

\bibitem[Kirby and Kronenberg, 1987]{Kirby:Kronenberg:1987}Kirby,
S. H. and A. K. Kronenberg (1987), Rheology of the lithosphere: Selected
topics, \textit{Reviews of Geophysics, 25}, 1219-1244.

\bibitem[Knopoff and Ni, 2001]{Knopoff:Ni:2001}Knopoff, L. and X.X.
Ni (2001), Numerical instability at the edge of a dynamic fracture,
\emph{Geophysical Journal International,}\textit{\emph{ }}\textit{147}(3),
1-6, doi: 10.1046/j.1365-246x.2001.01567.x.

\bibitem[Kojic and Bathe, 1987]{Kojic:Bathe:1987}Kojic, M. and K.-J.
Bathe (1987), The `Effective Stress-Function' Algorithm for Thermo-Elasto-Plasticity
and Creep, \emph{Int. J. Num. Meth. Eng}.\emph{, 24}, 1509-1532.

\bibitem[McGarr, 1988]{McGarr:1988}McGarr, A. (1988), On the state
of lithospheric stress in the absence of applied tectonic forces,
\textit{Journal of Geophysical Research}, \textit{93}, 13,609-13,617.

\bibitem[Menke, 1984]{Menke:1984}Menke, W. (1984), \textit{Geophysical
Data Analysis: Discrete Inverse Theory}, Academic Press, Inc., Orlando,
260 pp.

\bibitem[Okada, 1992]{Okada:1992}Okada, Y., Internal deformation
due to shear and tensile faults in a half-space (1992), \textit{Bull.
Seismol. Soc. Am.}, \textit{83}, 1018-1040.

\bibitem[Paterson, 1994]{Paterson:1994}Paterson, W. S. B. (1994),
\textit{The Physics of Glaciers, Third Edition}, Elsevier Science
Ltd., Oxford, 480 pp.

\bibitem[Prentice, 1968]{Prentice:1968}Prentice, J. H, (1968), Dimensional
problem of the power law in rheology, \textit{Nature}, \textit{217},
157.

\bibitem[Savage and Prescott, 1978]{Savage:Prescott:1978}Savage,
J. C. and W. H. Prescott (1978), Asthenosphere readjustment and the
earthquake cycle, \textit{Journal of Geophysical} \textit{Research}\emph{,
83}, 3369-3376.

\bibitem[Taylor, 2003]{Taylor:2003}Taylor, R.L. (2003), `FEAP--A
Finite Element Analysis Program', \textit{Version 7.5 Theory Manual},
154 pp.

\bibitem[Timoshenko and Goodier, 1987]{Timoshenko:Goodier:1987}Timoshenko,
S.P. and J.N. Goodier (1987), \textit{Theory of Elasticity, Third
Edition}, McGraw-Hill, New York, 567 pp.

\bibitem[Williams et al., 2005]{Williams:etal:2005}Williams, C.A.,
B. Aagaard, and M.G. Knepley (2005), Development of software for studying
earthquakes across multiple spatial and temporal scales by coupling
quasi-static and dynamic simulations, \emph{Eos Trans. AGU, 86}(52),
Fall Meet. Suppl., Abstract S53A-1072.

\bibitem[Williams, 2006]{Williams:2006}Williams, C.A. (2006), Development
of a package for modeling stress in the lithosphere, \emph{Eos Trans.
AGU,} \emph{87}(36), Jt. Assem. Suppl., Abstract T24A-01 Invited.

\bibitem[Zienkiewicz and Taylor, 2000]{Zienkiewicz:Taylor:2000}Zienkiewicz,
O.C. and R.L. Taylor (2000), \textit{The Finite Element Method, Fifth
Edition, Volume 2: Solid Mechanics}, Butterworth-Heinemann, Oxford,
459 pp.125\end{thebibliography}

\end{document}
