\chapter{Tutorials}

\section{Overview}

Each tutorial is a self-contained lesson in how to use PyLith. The
tutorials increase in degree of complexity from one to the next. 

\subsection{Prerequisites}

Before you begin any of the tutorials, you will need to install PyLith
following the instructions in section~\ref{sec:install}. In order to
work through a full tutorial, in addition to \application{PyLith} you
will need \href{http://www.hpfem.jku.at/netgen/}{\application{NETGEN}}
to generate the mesh and \href{http://www.paraview.org}{ParaView} to
view simulation results. You may use other packages, but some adaption
from what is described here will be necessary. Alternatively, you can
just complete a subset of the tutorial using files provided (as
described below), skipping the steps for which you do not have the
proper software packages installed.

The files needed to work through the tutorials have been collected
into a single tarball. Each tutorial contains instructions for
downloading and unpacking the tutorial tarball.

\subsection{Tutor}

Each tutorial includes a simple Python script, \filename{tutor.py}, to
help with miscellaneous tasks, such as copying provided files into a
working directory. This script can (1) check to make sure the files
necessary for a given step in the tutorial exist, (2) retrieve any
missing files from the tutorial archive directory that are needed for
a given step, or (3) prepare the work area for a given step by
removing old files that would otherwise be overwritten. You can run
\command{tutor.py} with the \option{-h} option for more
information. We will use this script at the beginning of each step of
the tutorials to retrieve files as necessary from the tutorial archive
directory.

\begin{tip}
  When retrieving files from the archive directory, \command{tutor.py}
  will not overwrite files that already exist in the workarea
  directory. This means that if you mangle files in the working area,
  you should remove them and let the tutor retrieve clean copies.
\end{tip}

\input{splittest/splitcube.tex}
\input{reversenog/reversenog.tex}
